---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ramen

<!-- badges: start -->
[![R-CMD-check](https://github.com/admarhi/ramen/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/admarhi/ramen/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/admarhi/ramen/branch/main/graph/badge.svg)](https://codecov.io/gh/admarhi/ramen?branch=main)
<!-- badges: end -->

**ramen** (**R**econstruction and **A**lignment of **M**icrobial
**E**xchange **N**etworks) compares microbial communities by their
metabolic exchange networks rather than by species composition. Given
species-metabolite-flux data (e.g. from community flux balance
analysis), ramen reconstructs directed metabolite-to-metabolite
networks and quantifies how functionally similar different communities
are -- even when they share no species at all.

## Quick start

```{r quick-start, message = FALSE}
library(ramen)
data("misosoup24")

## Build two consortium objects
cm1 <- ConsortiumMetabolism(
    misosoup24[[1]], name = names(misosoup24)[1],
    species_col = "species",
    metabolite_col = "metabolites",
    flux_col = "fluxes"
)
cm2 <- ConsortiumMetabolism(
    misosoup24[[2]], name = names(misosoup24)[2],
    species_col = "species",
    metabolite_col = "metabolites",
    flux_col = "fluxes"
)

## Align them
cma <- align(cm1, cm2)
scores(cma)
```

For the full workflow -- from data import through alignment,
functional groups, and visualization -- see
`vignette("ramen", package = "ramen")`.

## Installation

Install the development version from GitHub:

```{r install, eval = FALSE}
pak::pkg_install("admarhi/ramen")
# or
devtools::install_github("admarhi/ramen")
```

## Core workflow

1. **Import** species-metabolite-flux data (from MiSoSoup YAML,
   CSV, or any data.frame)
2. **Build** `ConsortiumMetabolism` objects (one per community)
3. **Group** them into a `ConsortiumMetabolismSet` (computes
   pairwise overlap and clustering)
4. **Align** communities pairwise or across the full set using
   five complementary metrics (FOS, Jaccard, Bray-Curtis,
   Redundancy Overlap, MAAS)
5. **Visualise** with heatmaps, network plots, dendrograms, and
   functional group analyses

All three classes inherit from Bioconductor's
`TreeSummarizedExperiment`, so standard accessor functions
(`assay()`, `colData()`, `dim()`, etc.) work out of the box.

## Vignettes

- `vignette("ramen")` -- full introduction and analysis pipeline
- `vignette("alignment")` -- alignment metrics, coverage ratios,
  and permutation p-values
- `vignette("visualisation")` -- gallery of all plot types
