---
title: "Alignment of Microbial Consortia"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Alignment of Microbial Consortia}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 5
)
```

# Introduction

The alignment system in `ramen` quantifies how similar two or more
microbial communities are from a functional perspective -- not by
which species are present, but by which metabolite-to-metabolite
pathways they catalyse. `align()` supports three modes: **pairwise**
(two consortia), **multiple** (all consortia in a set), and
**search** (one query consortium against every member of a set).
All three return a `ConsortiumMetabolismAlignment` (CMA) object
containing similarity scores and (depending on the mode) pathway
correspondences, a consensus network, or a ranked hit table.

This vignette covers the alignment system in depth. For a general
introduction to the package, see
`vignette("ramen", package = "ramen")`.

**Note on help lookup.** If `tibble` or `pillar` is loaded in your
session, `?align` may resolve to `pillar::align()` first. To view the
documentation for this package's generic, use `?ramen::align`.

```{r load-library, message = FALSE}
library(ramen)
```

# Mathematical formulation

This section gives the formal definitions of every quantity computed
by `align()`. Notation follows Hirt (2025), the master's thesis from
which the package originated; equation numbers prefixed `(thesis
eq. ...)` refer to that document.

## Notation

Let $\Omega = \{\omega^{(1)}, \dots, \omega^{(n)}\}$ denote a
collection of consortia (thesis eq. 1). For one consortium
$\omega^{(\alpha)}$, write $\mathcal{M}^{(\alpha)}$ for its set of
metabolites; for two consortia under comparison let
$\mathcal{M} = \mathcal{M}^{(\alpha)} \cup \mathcal{M}^{(\beta)}$
with $|\mathcal{M}| = m$ (thesis eq. 3). All assays are square
$m \times m$ matrices on this **union metabolite space**. We write
$X = \mathbf{X}^{\mathrm{bin},(\alpha)}$ and
$Y = \mathbf{X}^{\mathrm{bin},(\beta)}$ for the two binary assays
under comparison; analogous symbols $C_X, P_X, N_X$ denote the
Consumption, Production, and nSpecies assays. The
**$L_1$ norm** $\lVert A \rVert_1 = \sum_{i,j} |A_{ij}|$ is the sum
of all entries; $A \odot B$ is the elementwise (Hadamard) product.

A **pathway** in `ramen` is a directed
metabolite-to-metabolite pair $(i, j)$ such that some species in the
consortium can consume $i$ and some species (possibly the same one)
can produce $j$. Pathways are not biochemical reactions: they are
**capabilities** inferred from the Cartesian product of each species'
uptake set and secretion set. A consortium of two species, one
consuming $\{a, b\}$ and producing $\{c, d\}$, contributes the four
pathways $\{a \to c, a \to d, b \to c, b \to d\}$ even though only
two exchange reactions exist. See the *Cross-product inflation*
caveat below.

## Assay matrices

Each `ConsortiumMetabolism` stores six assays (thesis eq. 1, 2):

- **Binary** $\mathbf{X}^{(\alpha,1)}$: $X^{(\alpha,1)}_{ij} = 1$ if
  any species in $\omega^{(\alpha)}$ realises pathway $i \to j$,
  else $0$.
- **nSpecies** $\mathbf{X}^{(\alpha,2)}$: number of distinct species
  supporting $i \to j$.
- **Consumption** $\mathbf{X}^{(\alpha,3)}$:
  $F^{(c)}_{ij} = \sum_s f^{(c,s)}_{ij}$, the summed uptake flux of
  metabolite $i$ across all species participating in $i \to j$.
- **Production** $\mathbf{X}^{(\alpha,4)}$:
  $F^{(p)}_{ij} = \sum_s f^{(p,s)}_{ij}$, analogously for secretion.
- **EffectiveConsumption** $\mathbf{X}^{(\alpha,5)}$ and
  **EffectiveProduction** $\mathbf{X}^{(\alpha,6)}$: see below.

All six are constructed as the Cartesian product
$\mathrm{consumed}(s) \times \mathrm{produced}(s)$ for each species
$s$, then aggregated across species. This **cross-product** is the
source of the density inflation discussed under
*Limitations and caveats*.

## Effective flux (Hill-1 perplexity)

For a pathway $(i, j)$ with participating species $s_1, \dots, s_n$,
let $p^{(\phi,s)}_{ij} = f^{(\phi,s)}_{ij} / F^{(\phi)}_{ij}$ be the
fraction of total flux of type $\phi \in \{c, p\}$ contributed by
species $s$. The Shannon entropy of this distribution is
$H_{ij}^{(\phi)} = -\sum_s p^{(\phi,s)}_{ij}
\log_2 p^{(\phi,s)}_{ij}$, and the **Hill-1 perplexity** is
$\,^{1}\!D_{ij}^{(\phi)} = 2^{H_{ij}^{(\phi)}}$. This is the
*effective number of species* contributing to the pathway. The
EffectiveConsumption / EffectiveProduction assays store
$$
\mathbf{X}^{(\alpha,5)}_{ij} = \mathrm{round}\bigl(
2^{H^{(c)}_{ij}}, 2 \bigr),
\qquad
\mathbf{X}^{(\alpha,6)}_{ij} = \mathrm{round}\bigl(
2^{H^{(p)}_{ij}}, 2 \bigr).
$$
Note that the implementation stores the **perplexity alone**, not
the entropy-weighted flux $F^{(\phi)}_{ij} \cdot 2^{H}$ given as
thesis eq. 2; the package follows the implementation. See the
*Hill-1 saturation* caveat for behaviour on small consortia.

## Pairwise alignment metrics

Throughout, $X$ and $Y$ are binary assays expanded to the union
metabolite space, and $C_X, P_X, N_X$ are the Consumption,
Production, and nSpecies assays for consortium $X$.

### FOS (Szymkiewicz-Simpson overlap)

$$
S_{\mathrm{FOS}}(X, Y)
= \frac{\lVert X \odot Y \rVert_1}{\min\bigl(
\lVert X \rVert_1, \lVert Y \rVert_1 \bigr)}
\qquad \text{(thesis eq. 5)}.
$$
Range $[0, 1]$. **Asymmetric** in size: $S_{\mathrm{FOS}} = 1$
whenever the smaller binary network is a subset of the larger,
which is what motivates the coverage ratios below. Returns $0$ by
convention if either matrix is empty. This is the default `method`.

### Jaccard

$$
S_{\mathrm{J}}(X, Y)
= \frac{\lVert X \odot Y \rVert_1}{\lVert X \rVert_1
+ \lVert Y \rVert_1 - \lVert X \odot Y \rVert_1}.
$$
Range $[0, 1]$. **Symmetric**. For binary $X, Y$ this is the
classical $|S_X \cap S_Y| / |S_X \cup S_Y|$ on pathway sets.

### Bray-Curtis (similarity, on combined fluxes)

$$
S_{\mathrm{BC}}(X, Y)
= \max\!\Bigl(0,\; 1 - \frac{
\lVert |C_X - C_Y| \rVert_1 + \lVert |P_X - P_Y| \rVert_1
}{\lVert C_X \rVert_1 + \lVert C_Y \rVert_1
+ \lVert P_X \rVert_1 + \lVert P_Y \rVert_1}\Bigr).
$$
The implementation operates on the **combined Consumption +
Production magnitudes** treated as one abundance vector; direction
is conflated. Range $[0, 1]$, with $1$ at identical fluxes. The
$\max(0, \cdot)$ clamp guards against floating-point near-negative
zero; this is the consequence of the B7 fix and is why
`brayCurtis = NA` is returned for unweighted CMs (the metric is
undefined when fluxes are unit placeholders). See the
`?ramen::align` `@note`.

### RedundancyOverlap (weighted Jaccard / Ruzicka)

Using $\min(a, b) = (a + b - |a - b|)/2$ and
$\max(a, b) = (a + b + |a - b|)/2$,
$$
S_{\mathrm{RO}}(X, Y)
= \frac{\sum_{i,j} \min\bigl(N_{X,ij}, N_{Y,ij}\bigr)}{
\sum_{i,j} \max\bigl(N_{X,ij}, N_{Y,ij}\bigr)}.
$$
Range $[0, 1]$. **Collapse to Jaccard at $N = 1$.** When all
positive entries of $N_X$ and $N_Y$ equal $1$ (every supported
pathway is single-species) the numerator becomes the indicator
intersection $\lVert X \odot Y \rVert_1$ and the denominator
becomes the indicator union $\lVert X \rVert_1 + \lVert Y \rVert_1
- \lVert X \odot Y \rVert_1$, which is exactly $S_{\mathrm{J}}$.
Small communities therefore yield identical Jaccard and
RedundancyOverlap scores.

### MAAS (Metabolic Alignment Aggregate Score)

A weighted convex combination of the four base metrics. Let
$\mathcal{A}$ be the set of metrics whose inputs are available
(Bray-Curtis and RedundancyOverlap drop out for unweighted CMs).
$$
S_{\mathrm{MAAS}}(X, Y)
= \sum_{m \in \mathcal{A}} \frac{w_m}{\sum_{m' \in \mathcal{A}}
w_{m'}} \cdot S_m(X, Y).
$$
Default weights are
$(w_{\mathrm{FOS}}, w_{\mathrm{J}}, w_{\mathrm{BC}},
w_{\mathrm{RO}}) = (0.4, 0.2, 0.2, 0.2)$. **Disclosure:** these
defaults are not derived from data; they encode a soft preference
for FOS as the primary metric and were not tuned by sensitivity
analysis. A formal sensitivity sweep is deferred to the methods
paper.

### Coverage ratios

To disambiguate FOS = 1 from full functional equivalence,
$$
\mathrm{cov}_{\mathrm{query}}
= \frac{\lVert X \odot Y \rVert_1}{\lVert X \rVert_1},
\qquad
\mathrm{cov}_{\mathrm{ref}}
= \frac{\lVert X \odot Y \rVert_1}{\lVert Y \rVert_1}.
$$
$S_{\mathrm{FOS}} = 1$ together with low $\mathrm{cov}_{\mathrm{ref}}$
indicates that the query is a strict subset of the reference, not
a functional twin.

## Permutation null model

For a metric $S$ and observed value $s_{\mathrm{obs}} = S(X, Y)$, a
$p$-value is computed by **degree-preserving rewire** of the query
binary network, treated as a directed graph
$G_X = (\mathcal{M}, E_X)$. Let $\mathrm{rewire}(G_X)$ denote the
graph obtained from $G_X$ by a sequence of edge swaps that preserve
the in-degree and out-degree sequence; `igraph::keeping_degseq` is
used with $\mathtt{niter} = 10 \cdot |E_X|$. The permutation
$p$-value is
$$
p(s_{\mathrm{obs}})
= \frac{1 + \sum_{k=1}^{B} \mathbf{1}\bigl[
S(\mathrm{rewire}_k(G_X), Y) \ge s_{\mathrm{obs}}\bigr]}{1 + B},
$$
with $B = \mathtt{nPermutations}$ (default $999$). The null
hypothesis is *"the observed similarity is no greater than expected
under random rewiring of $X$ that preserves each metabolite's
participation count"*.

**Limitation (methods-paper concern, not a current correctness
issue).** Degree-preserving rewire is the conventional null in
ecological network analysis but is **not specifically motivated for
FBA-derived networks**: edge presence in $X$ reflects mass-balance
constraints on the underlying genome-scale metabolic models, plus
the solver's choice among alternate optima. A null that respects
those constraints (for example a species-set permutation, or a
randomised-abundance null in the MICOM sense) would be more
defensible. The current null is retained for reproducibility with
prior `ramen` versions; sensitivity to null-model choice is on the
methods-paper agenda.

## When to use what: alignment vs functional groups

`ramen` exposes two superficially similar comparisons. They answer
**different questions** and are not interchangeable.

**Alignment** (`align(CM, CM)`, `align(CMS)`, `align(CM, CMS)`)
compares two or more *named consortia* by their metabolic exchange
networks. The unit of analysis is the consortium; the question is
*"how similar are these communities by what they collectively do?"*

**Functional groups** (`functionalGroups()`) cluster *species* by
the Jaccard similarity of their per-species pathway sets, either
within a single consortium or pooled across a CMS. The unit of
analysis is the species; the question is *"which species play
equivalent metabolic roles?"* See `vignette("ramen", package =
"ramen")` for functional groups.

## Limitations and caveats

The metric formulas above are not a license to run alignment on
arbitrary inputs. Several known issues affect interpretation.

**Cross-product inflation.** Pathways are inferred from the Cartesian
product of each species' uptake and secretion sets, not from
biochemical reactions. For a consortium of two species
$S_1 : \{a, b\} \to \{c, d\}$ and $S_2 : \{c, d\} \to \{a, b\}$,
the Binary assay has $|\{a, b\} \times \{c, d\}|
+ |\{c, d\} \times \{a, b\}| = 8$ nonzero entries in a $4 \times 4$
matrix, giving density $0.5$ from only $4$ exchange reactions. This
is a **capability-level** representation, not a mechanism-level one,
and the inflation grows with $\min(|\mathrm{cons}|, |\mathrm{prod}|)$
per species. Compare consortia of similar size and composition; do
not interpret raw Binary density as a mechanistic property.

**Hill-1 saturation on small consortia.** EffectiveConsumption /
EffectiveProduction store Hill-1 perplexity rounded to two decimals.
For two-species consortia the perplexity rarely exceeds $\approx 1.05$
(one species typically dominates the flux at any given pathway), and
the `round(., 2)` step collapses near-degenerate distributions to
exactly $1.0$ -- indistinguishable from true single-species
pathways. Effective C/P is informative only on consortia where
several species contribute comparable flux to the same pathway.

**Flux reversibility and no binarisation threshold.**
`ConsortiumMetabolism()` admits any nonzero flux: a flux of $10^{-12}$
produces an edge identical to a flux of $10^3$. FBA solvers
(MiSoSoup, MICOM, cobrapy) routinely produce alternate optima that
differ only in sign of small fluxes near the noise floor; under the
current convention this turns a "consume $a$, produce $b$" pathway
into "consume $b$, produce $a$", which the Binary assay treats as a
different pathway. No `flux_tolerance` argument is currently exposed.
For solver outputs we recommend pre-filtering by solver tolerance
(typically $10^{-9}$ for COBRA / cobrapy) before constructing CMs.

**Bray-Curtis on unweighted CMs returns NA.** The fix for the
floating-point near-negative-zero bug (B7) introduced an explicit
$\max(0, \cdot)$ clamp that propagates `NA` when fluxes are unit
placeholders rather than measured values. This is documented in the
`align()` `@note`. To get a non-`NA` Bray-Curtis, build CMs from
non-unit fluxes.

**Alternative optima are not biological signal.** MiSoSoup is a MILP
enumerator: for a fixed metabolic model it returns multiple
alternative optimal solutions to the same growth problem. Such
alternatives typically share most pathways and produce very high
pairwise FOS values that reflect **solver consistency, not
biological similarity**. The same caveat applies to outputs from
**MICOM** and **cobrapy**, where the choice of solver, tolerance,
and tie-breaking affects which exchange fluxes appear in the
solution. When interpreting overlap, compare consortia derived from
*different* models, conditions, or experimental contexts; treat
overlap between alternates of one model as a sanity check on the
enumeration, not as ecology.

**Minimum production flux drives the alignment.** Because pathways
are inferred from the existence of any nonzero production flux for
a given species and metabolite, low-magnitude secretions (whether
biologically real or solver noise) determine the topology of the
Binary assay just as much as high-magnitude ones. This is a design
decision, not a bug: it preserves the **set of metabolic
capabilities** the consortium can realise. Users who want to weight
by magnitude should use Bray-Curtis or RedundancyOverlap rather than
FOS / Jaccard, and should consider thresholding fluxes upstream of
`ConsortiumMetabolism()`.

# Test data

We build six consortia from the bundled `misosoup24` dataset to use
throughout this vignette.

```{r create-data, message = FALSE}
data("misosoup24")
cm_list <- lapply(seq_len(6), function(i) {
    ConsortiumMetabolism(
        misosoup24[[i]],
        name = names(misosoup24)[i]
    )
})

cms <- ConsortiumMetabolismSet(cm_list, name = "Demo")
```

# Pairwise alignment

## Basic usage

`align(CM, CM)` compares two `ConsortiumMetabolism` objects and
returns a CMA with `Type = "pairwise"`.

```{r pairwise-basic}
cma <- align(cm_list[[1]], cm_list[[2]])
cma
```

## Similarity metrics

Five metrics are available via the `method` argument. Regardless of
which is selected as the primary score, all applicable metrics are
always computed and stored.

```{r pairwise-all-methods}
cma_fos <- align(cm_list[[1]], cm_list[[2]], method = "FOS")
cma_jac <- align(cm_list[[1]], cm_list[[2]], method = "jaccard")
scores(cma_fos)
```

| Metric | Description |
|--------|-------------|
| FOS | Szymkiewicz-Simpson on binary matrices |
| Jaccard | Symmetric set similarity |
| Bray-Curtis | Flux-weighted similarity |
| Redundancy Overlap | Weighted Jaccard on nSpecies |
| MAAS | 0.4 FOS + 0.2 each of the rest |

The **Metabolic Alignment Aggregate Score** (MAAS) combines all
four metrics. Weights are renormalized when some metrics are
unavailable (e.g., unweighted networks lack Bray-Curtis):

```{r pairwise-maas}
cma_maas <- align(cm_list[[1]], cm_list[[2]], method = "MAAS")
scores(cma_maas)
```

## The FOS subset property

FOS uses the Szymkiewicz-Simpson coefficient, which divides by the
**smaller** network. This means FOS = 1 whenever a small consortium
is a strict functional subset of a larger one -- even if the larger
consortium has many additional pathways.

To detect this, `ramen` reports **coverage ratios** alongside the
similarity metrics:

```{r coverage-ratios}
scores(cma_fos)[c("FOS", "coverageQuery", "coverageReference")]
```

- `coverageQuery`: fraction of the query's pathways found in the
    reference
- `coverageReference`: fraction of the reference's pathways found
    in the query

When FOS is high but one coverage ratio is low, the alignment
represents a subset relationship rather than true functional
equivalence. For symmetric similarity, use Jaccard instead.

## Pathway correspondences

The alignment classifies every metabolite-to-metabolite pathway as
shared, unique to the query, or unique to the reference.

```{r pairwise-pathways}
## All pathways in the alignment
head(pathways(cma))

## Shared pathways only
shared <- pathways(cma, type = "shared")
nrow(shared)

## Unique pathways (returns a list with $query and $reference)
unique_pw <- pathways(cma, type = "unique")
nrow(unique_pw$query)
nrow(unique_pw$reference)
```

## Permutation p-values

Statistical significance is assessed by degree-preserving network
rewiring. The query network's pathways are shuffled while preserving
each metabolite's degree, and the metric is recomputed under the
null distribution. For MAAS, only the binary network topology is
permuted; the weighted assays (Consumption, Production, nSpecies)
remain fixed, so the null distribution reflects topological
variation in the composite score.

```{r pairwise-pvalue}
cma_p <- align(
    cm_list[[1]],
    cm_list[[2]],
    method = "FOS",
    computePvalue = TRUE,
    nPermutations = 99L
)
scores(cma_p)
```

## Visualization

### Network plot

The network view shows shared (green), query-unique (blue), and
reference-unique (red) pathways as a directed metabolite flow
graph.

```{r plot-network, fig.cap = "Pairwise alignment network."}
plot(cma, type = "network")
```

### Score bar chart

```{r plot-scores-pair, fig.cap = "Pairwise metric scores."}
plot(cma, type = "scores")
```

# Multiple alignment

## Aligning a consortium set

`align(CMS)` computes pairwise similarities across all consortia
in a `ConsortiumMetabolismSet` and returns a CMA with
`Type = "multiple"`.

```{r multiple-create, message = FALSE}
cma_mult <- align(cms)
cma_mult
```

## Similarity matrix

The similarity matrix is an n x n symmetric matrix with 1s on the
diagonal. For FOS, this is derived from the pre-computed CMS
overlap matrix:

```{r multiple-simmat}
round(similarityMatrix(cma_mult), 3)
```

## Summary scores

For a multiple alignment, the primary score is the **median** of
all pairwise scores. `scores()` returns full summary statistics:

```{r multiple-scores}
scores(cma_mult)
```

## Consensus network and prevalence

Pathway prevalence counts how many consortia share each
metabolite-to-metabolite pathway. This enables classification of
pathways as core (present in most consortia) or niche (present in
few).

```{r multiple-prevalence}
prev <- prevalence(cma_mult)
head(prev[order(-prev$nConsortia), ])

## Distribution of prevalence
table(prev$nConsortia)
```

The `pathways()` method with `type = "consensus"` returns the same
information:

```{r multiple-consensus}
head(pathways(cma_mult, type = "consensus"))
```

## Visualization

### Heatmap

The heatmap shows pairwise similarities with dendrogram-based
ordering:

```{r plot-heatmap, fig.cap = "Similarity heatmap."}
plot(cma_mult, type = "heatmap")
```

### Score summary

```{r plot-scores-mult, fig.cap = "Multiple alignment summary scores."}
plot(cma_mult, type = "scores")
```

# Database search

`align(CM, CMS)` compares a single query consortium against every
member of a set and returns a CMA with `Type = "search"`. This is
the natural call for questions like "which consortium in my
database is most functionally similar to this query?"

## Basic search

We hold out the first consortium as a query and search it against a
database built from the remaining five:

```{r search-setup, message = FALSE}
query <- cm_list[[1]]
db <- ConsortiumMetabolismSet(cm_list[-1], name = "db")
```

```{r search-basic}
hits <- align(query, db)
hits
```

The top-hit name and score sit on `ReferenceName` / `PrimaryScore`.
The full ranked table is stored in `Scores$ranking`:

```{r search-ranking}
ranking <- scores(hits)$ranking
head(ranking)
```

Each row holds the reference consortium's name, the primary `score`
(under the requested `method`), all four individual metric columns,
and the coverage ratios against the query. Rows are pre-sorted by
`score` in descending order.

## Top-K hits

Use `topK` to truncate the ranked table -- useful when the database
is large and only the best matches matter:

```{r search-topk}
hits_top3 <- align(query, db, topK = 3L)
scores(hits_top3)$ranking
```

`similarityMatrix()` returns a 1 x n row vector (or 1 x topK if
truncated), with the query name as the row label:

```{r search-simmat}
round(similarityMatrix(hits_top3), 3)
```

Pathway correspondences always reflect the single overall top hit,
regardless of `topK`:

```{r search-correspondences}
head(pathways(hits_top3, type = "shared"))
```

## Choosing metrics

By default all four base metrics are computed for every database
member. For large databases, restricting `metrics` skips the
weighted-assay expansion and can be substantially faster:

```{r search-metrics}
hits_fos <- align(query, db, metrics = "FOS")
scores(hits_fos)$ranking[,
    c("reference", "score", "brayCurtis")
]
```

Columns for skipped metrics remain in the ranking table but are
filled with `NA`, making the schema stable across calls.

## Significance of the top hit

As in pairwise alignment, `computePvalue = TRUE` runs a
degree-preserving permutation test. For database search, the test
is applied only to the top hit -- a BLAST-style convention that
keeps the cost comparable to a single pairwise p-value:

```{r search-pvalue}
hits_p <- align(
    query,
    db,
    computePvalue = TRUE,
    nPermutations = 99L
)
hits_p
```

The `show()` output reports the top hit, its score, and the p-value
directly; `scores(hits_p)` gives the full numeric breakdown including
`pvalue`.

If statistical confidence is needed for several top hits, re-run
`align()` in pairwise mode against each candidate individually.

# Accessor reference

| Accessor | Alignment type | Returns |
|----------|---------------|---------|
| `scores()` | all | Named list of scores (+ `$ranking` for search) |
| `pathways()` | all | data.frame of all pathways |
| `pathways(type = "shared")` | pairwise, search | shared pathways |
| `pathways(type = "unique")` | pairwise, search | list(query, reference) |
| `pathways(type = "consensus")` | multiple | data.frame with prevalence |
| `similarityMatrix()` | multiple, search | n x n or 1 x n numeric matrix |
| `prevalence()` | multiple | data.frame with nConsortia |
| `metabolites()` | all | Character vector of metabolites |

Type guards prevent misuse -- for example, calling
`pathways(type = "consensus")` on a pairwise alignment raises an
informative error.

# Session info

```{r session-info}
sessionInfo()
```
