---
title: "Alignment of Microbial Consortia"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Alignment of Microbial Consortia}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 5
)
```

# Introduction

The alignment system in `ramen` quantifies how similar two or more
microbial communities are from a functional perspective -- not by
which species are present, but by which metabolite-to-metabolite
pathways they catalyse. The result is a
`ConsortiumMetabolismAlignment` (CMA) object containing similarity
scores, pathway correspondences, and (for multiple alignments) a
consensus network.

This vignette covers the alignment system in depth. For a general
introduction to the package, see
`vignette("ramen", package = "ramen")`.

```{r load-library, message = FALSE}
library(ramen)
```

# Test data

We build six consortia from the bundled `misosoup24` dataset to use
throughout this vignette.

```{r create-data, message = FALSE}
data("misosoup24")
cm_list <- lapply(seq_len(6), function(i) {
    ConsortiumMetabolism(
        misosoup24[[i]],
        name = names(misosoup24)[i],
        species_col = "species",
        metabolite_col = "metabolites",
        flux_col = "fluxes"
    )
})

cms <- ConsortiumMetabolismSet(cm_list, name = "Demo")
```

# Pairwise alignment

## Basic usage

`align(CM, CM)` compares two `ConsortiumMetabolism` objects and
returns a CMA with `Type = "pairwise"`.

```{r pairwise-basic}
cma <- align(cm_list[[1]], cm_list[[2]])
cma
```

## Similarity metrics

Five metrics are available via the `method` argument. Regardless of
which is selected as the primary score, all applicable metrics are
always computed and stored.

```{r pairwise-all-methods}
cma_fos <- align(cm_list[[1]], cm_list[[2]], method = "FOS")
cma_jac <- align(cm_list[[1]], cm_list[[2]], method = "jaccard")
scores(cma_fos)
```

| Metric | Description |
|--------|-------------|
| FOS | Szymkiewicz-Simpson on binary matrices |
| Jaccard | Symmetric set similarity |
| Bray-Curtis | Flux-weighted similarity |
| Redundancy Overlap | Weighted Jaccard on nSpecies |
| MAAS | 0.4 FOS + 0.2 each of the rest |

The **Metabolic Alignment Aggregate Score** (MAAS) combines all
four metrics. Weights are renormalized when some metrics are
unavailable (e.g., unweighted networks lack Bray-Curtis):

```{r pairwise-maas}
cma_maas <- align(cm_list[[1]], cm_list[[2]], method = "MAAS")
scores(cma_maas)
```

## Pathway correspondences

The alignment classifies every metabolite-to-metabolite pathway as
shared, unique to the query, or unique to the reference.

```{r pairwise-pathways}
## All pathways in the alignment
head(pathways(cma))

## Shared pathways only
shared <- pathways(cma, type = "shared")
nrow(shared)

## Unique pathways (returns a list with $query and $reference)
unique_pw <- pathways(cma, type = "unique")
nrow(unique_pw$query)
nrow(unique_pw$reference)
```

## Permutation p-values

Statistical significance is assessed by degree-preserving network
rewiring. The query network's pathways are shuffled while preserving
each metabolite's degree, and the metric is recomputed under the
null distribution.

```{r pairwise-pvalue}
cma_p <- align(
    cm_list[[1]],
    cm_list[[2]],
    method = "FOS",
    computePvalue = TRUE,
    nPermutations = 99L
)
scores(cma_p)
```

## Visualization

### Network plot

The network view shows shared (green), query-unique (blue), and
reference-unique (red) pathways as a directed metabolite flow
graph.

```{r plot-network, fig.cap = "Pairwise alignment network."}
plot(cma, type = "network")
```

### Score bar chart

```{r plot-scores-pair, fig.cap = "Pairwise metric scores."}
plot(cma, type = "scores")
```

# Multiple alignment

## Aligning a consortium set

`align(CMS)` computes pairwise similarities across all consortia
in a `ConsortiumMetabolismSet` and returns a CMA with
`Type = "multiple"`.

```{r multiple-create, message = FALSE}
cma_mult <- align(cms)
cma_mult
```

## Similarity matrix

The similarity matrix is an n x n symmetric matrix with 1s on the
diagonal. For FOS, this is derived from the pre-computed CMS
overlap matrix:

```{r multiple-simmat}
round(similarityMatrix(cma_mult), 3)
```

## Summary scores

For a multiple alignment, the primary score is the **median** of
all pairwise scores. `scores()` returns full summary statistics:

```{r multiple-scores}
scores(cma_mult)
```

## Consensus network and prevalence

Pathway prevalence counts how many consortia share each
metabolite-to-metabolite pathway. This enables classification of
pathways as core (present in most consortia) or niche (present in
few).

```{r multiple-prevalence}
prev <- prevalence(cma_mult)
head(prev[order(-prev$nConsortia), ])

## Distribution of prevalence
table(prev$nConsortia)
```

The `pathways()` method with `type = "consensus"` returns the same
information:

```{r multiple-consensus}
head(pathways(cma_mult, type = "consensus"))
```

## Visualization

### Heatmap

The heatmap shows pairwise similarities with dendrogram-based
ordering:

```{r plot-heatmap, fig.cap = "Similarity heatmap."}
plot(cma_mult, type = "heatmap")
```

### Score summary

```{r plot-scores-mult, fig.cap = "Multiple alignment summary scores."}
plot(cma_mult, type = "scores")
```

# Accessor reference

| Accessor | Alignment type | Returns |
|----------|---------------|---------|
| `scores()` | both | Named list of scores |
| `pathways()` | both | data.frame of all pathways |
| `pathways(type = "shared")` | pairwise | data.frame of shared pathways |
| `pathways(type = "unique")` | pairwise | list(query, reference) |
| `pathways(type = "consensus")` | multiple | data.frame with prevalence |
| `similarityMatrix()` | multiple | n x n numeric matrix |
| `prevalence()` | multiple | data.frame with nConsortia |
| `metabolites()` | both | Character vector of metabolites |

Type guards prevent misuse -- for example, calling
`pathways(type = "shared")` on a multiple alignment raises an
informative error.

# Session info

```{r session-info}
sessionInfo()
```
