---
title: "Alignment of Microbial Consortia"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Alignment of Microbial Consortia}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 5
)
```

# Introduction

The alignment system in `ramen` quantifies how similar two or more
microbial communities are from a functional perspective -- not by
which species are present, but by which metabolite-to-metabolite
pathways they catalyse. `align()` supports three modes: **pairwise**
(two consortia), **multiple** (all consortia in a set), and
**search** (one query consortium against every member of a set).
All three return a `ConsortiumMetabolismAlignment` (CMA) object
containing similarity scores and (depending on the mode) pathway
correspondences, a consensus network, or a ranked hit table.

This vignette covers the alignment system in depth. For a general
introduction to the package, see
`vignette("ramen", package = "ramen")`.

```{r load-library, message = FALSE}
library(ramen)
```

# Test data

We build six consortia from the bundled `misosoup24` dataset to use
throughout this vignette.

```{r create-data, message = FALSE}
data("misosoup24")
cm_list <- lapply(seq_len(6), function(i) {
    ConsortiumMetabolism(
        misosoup24[[i]],
        name = names(misosoup24)[i]
    )
})

cms <- ConsortiumMetabolismSet(cm_list, name = "Demo")
```

# Pairwise alignment

## Basic usage

`align(CM, CM)` compares two `ConsortiumMetabolism` objects and
returns a CMA with `Type = "pairwise"`.

```{r pairwise-basic}
cma <- align(cm_list[[1]], cm_list[[2]])
cma
```

## Similarity metrics

Five metrics are available via the `method` argument. Regardless of
which is selected as the primary score, all applicable metrics are
always computed and stored.

```{r pairwise-all-methods}
cma_fos <- align(cm_list[[1]], cm_list[[2]], method = "FOS")
cma_jac <- align(cm_list[[1]], cm_list[[2]], method = "jaccard")
scores(cma_fos)
```

| Metric | Description |
|--------|-------------|
| FOS | Szymkiewicz-Simpson on binary matrices |
| Jaccard | Symmetric set similarity |
| Bray-Curtis | Flux-weighted similarity |
| Redundancy Overlap | Weighted Jaccard on nSpecies |
| MAAS | 0.4 FOS + 0.2 each of the rest |

The **Metabolic Alignment Aggregate Score** (MAAS) combines all
four metrics. Weights are renormalized when some metrics are
unavailable (e.g., unweighted networks lack Bray-Curtis):

```{r pairwise-maas}
cma_maas <- align(cm_list[[1]], cm_list[[2]], method = "MAAS")
scores(cma_maas)
```

## The FOS subset property

FOS uses the Szymkiewicz-Simpson coefficient, which divides by the
**smaller** network. This means FOS = 1 whenever a small consortium
is a strict functional subset of a larger one -- even if the larger
consortium has many additional pathways.

To detect this, `ramen` reports **coverage ratios** alongside the
similarity metrics:

```{r coverage-ratios}
scores(cma_fos)[c("FOS", "coverageQuery", "coverageReference")]
```

- `coverageQuery`: fraction of the query's pathways found in the
  reference
- `coverageReference`: fraction of the reference's pathways found
  in the query

When FOS is high but one coverage ratio is low, the alignment
represents a subset relationship rather than true functional
equivalence. For symmetric similarity, use Jaccard instead.

## Pathway correspondences

The alignment classifies every metabolite-to-metabolite pathway as
shared, unique to the query, or unique to the reference.

```{r pairwise-pathways}
## All pathways in the alignment
head(pathways(cma))

## Shared pathways only
shared <- pathways(cma, type = "shared")
nrow(shared)

## Unique pathways (returns a list with $query and $reference)
unique_pw <- pathways(cma, type = "unique")
nrow(unique_pw$query)
nrow(unique_pw$reference)
```

## Permutation p-values

Statistical significance is assessed by degree-preserving network
rewiring. The query network's pathways are shuffled while preserving
each metabolite's degree, and the metric is recomputed under the
null distribution. For MAAS, only the binary network topology is
permuted; the weighted assays (Consumption, Production, nSpecies)
remain fixed, so the null distribution reflects topological
variation in the composite score.

```{r pairwise-pvalue}
cma_p <- align(
    cm_list[[1]],
    cm_list[[2]],
    method = "FOS",
    computePvalue = TRUE,
    nPermutations = 99L
)
scores(cma_p)
```

## Visualization

### Network plot

The network view shows shared (green), query-unique (blue), and
reference-unique (red) pathways as a directed metabolite flow
graph.

```{r plot-network, fig.cap = "Pairwise alignment network."}
plot(cma, type = "network")
```

### Score bar chart

```{r plot-scores-pair, fig.cap = "Pairwise metric scores."}
plot(cma, type = "scores")
```

# Multiple alignment

## Aligning a consortium set

`align(CMS)` computes pairwise similarities across all consortia
in a `ConsortiumMetabolismSet` and returns a CMA with
`Type = "multiple"`.

```{r multiple-create, message = FALSE}
cma_mult <- align(cms)
cma_mult
```

## Similarity matrix

The similarity matrix is an n x n symmetric matrix with 1s on the
diagonal. For FOS, this is derived from the pre-computed CMS
overlap matrix:

```{r multiple-simmat}
round(similarityMatrix(cma_mult), 3)
```

## Summary scores

For a multiple alignment, the primary score is the **median** of
all pairwise scores. `scores()` returns full summary statistics:

```{r multiple-scores}
scores(cma_mult)
```

## Consensus network and prevalence

Pathway prevalence counts how many consortia share each
metabolite-to-metabolite pathway. This enables classification of
pathways as core (present in most consortia) or niche (present in
few).

```{r multiple-prevalence}
prev <- prevalence(cma_mult)
head(prev[order(-prev$nConsortia), ])

## Distribution of prevalence
table(prev$nConsortia)
```

The `pathways()` method with `type = "consensus"` returns the same
information:

```{r multiple-consensus}
head(pathways(cma_mult, type = "consensus"))
```

## Visualization

### Heatmap

The heatmap shows pairwise similarities with dendrogram-based
ordering:

```{r plot-heatmap, fig.cap = "Similarity heatmap."}
plot(cma_mult, type = "heatmap")
```

### Score summary

```{r plot-scores-mult, fig.cap = "Multiple alignment summary scores."}
plot(cma_mult, type = "scores")
```

# Database search

`align(CM, CMS)` compares a single query consortium against every
member of a set and returns a CMA with `Type = "search"`. This is
the natural call for questions like "which consortium in my
database is most functionally similar to this query?"

## Basic search

We hold out the first consortium as a query and search it against a
database built from the remaining five:

```{r search-setup, message = FALSE}
query <- cm_list[[1]]
db <- ConsortiumMetabolismSet(cm_list[-1], name = "db")
```

```{r search-basic}
hits <- align(query, db)
hits
```

The top-hit name and score sit on `ReferenceName` / `PrimaryScore`.
The full ranked table is stored in `Scores$ranking`:

```{r search-ranking}
ranking <- scores(hits)$ranking
head(ranking)
```

Each row holds the reference consortium's name, the primary `score`
(under the requested `method`), all four individual metric columns,
and the coverage ratios against the query. Rows are pre-sorted by
`score` in descending order.

## Top-K hits

Use `topK` to truncate the ranked table -- useful when the database
is large and only the best matches matter:

```{r search-topk}
hits_top3 <- align(query, db, topK = 3L)
scores(hits_top3)$ranking
```

`similarityMatrix()` returns a 1 x n row vector (or 1 x topK if
truncated), with the query name as the row label:

```{r search-simmat}
round(similarityMatrix(hits_top3), 3)
```

Pathway correspondences always reflect the single overall top hit,
regardless of `topK`:

```{r search-correspondences}
head(pathways(hits_top3, type = "shared"))
```

## Choosing metrics

By default all four base metrics are computed for every database
member. For large databases, restricting `metrics` skips the
weighted-assay expansion and can be substantially faster:

```{r search-metrics}
hits_fos <- align(query, db, metrics = "FOS")
scores(hits_fos)$ranking[,
    c("reference", "score", "brayCurtis")
]
```

Columns for skipped metrics remain in the ranking table but are
filled with `NA`, making the schema stable across calls.

## Significance of the top hit

As in pairwise alignment, `computePvalue = TRUE` runs a
degree-preserving permutation test. For database search, the test
is applied only to the top hit -- a BLAST-style convention that
keeps the cost comparable to a single pairwise p-value:

```{r search-pvalue}
hits_p <- align(
    query,
    db,
    computePvalue = TRUE,
    nPermutations = 99L
)
hits_p
```

The `show()` output reports the top hit, its score, and the p-value
directly; `scores(hits_p)` gives the full numeric breakdown including
`pvalue`.

If statistical confidence is needed for several top hits, re-run
`align()` in pairwise mode against each candidate individually.

# Accessor reference

| Accessor | Alignment type | Returns |
|----------|---------------|---------|
| `scores()` | all | Named list of scores (+ `$ranking` for search) |
| `pathways()` | all | data.frame of all pathways |
| `pathways(type = "shared")` | pairwise, search | shared pathways (top hit for search) |
| `pathways(type = "unique")` | pairwise, search | list(query, reference) |
| `pathways(type = "consensus")` | multiple | data.frame with prevalence |
| `similarityMatrix()` | multiple, search | n x n or 1 x n numeric matrix |
| `prevalence()` | multiple | data.frame with nConsortia |
| `metabolites()` | all | Character vector of metabolites |

Type guards prevent misuse -- for example, calling
`pathways(type = "consensus")` on a pairwise alignment raises an
informative error.

# Session info

```{r session-info}
sessionInfo()
```
