---
title: "The ConsortiumMetabolismSet Class"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{The ConsortiumMetabolismSet Class}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

# Introduction

The `ConsortiumMetabolismSet` (CMS) class groups multiple
`ConsortiumMetabolism` objects into a single analysis unit. When a CMS
is constructed, `ramen` automatically:

1. Expands all binary matrices to a shared **universal metabolite
   space**
2. Computes **pairwise overlap scores** (Functional Overlap Score)
   between every pair of consortia
3. Builds a **dendrogram** via hierarchical clustering on the
   dissimilarity matrix (1 - overlap)

This vignette covers CMS construction, accessors, the overlap matrix,
dendrogram operations, species and pathway classification, and cluster
extraction.

```{r load-library, message = FALSE}
library(ramen)
```

# Construction

## From a list of CMs

The `ConsortiumMetabolismSet()` constructor accepts any number of
`ConsortiumMetabolism` objects (or a list of them), plus a `name`.

```{r construct-synCM, message = FALSE}
cm_alpha <- synCM("Alpha", n_species = 5, max_met = 10, seed = 42)
cm_beta <- synCM("Beta", n_species = 5, max_met = 10, seed = 43)
cm_gamma <- synCM("Gamma", n_species = 4, max_met = 8, seed = 44)
cm_delta <- synCM("Delta", n_species = 6, max_met = 10, seed = 45)

cms_syn <- ConsortiumMetabolismSet(
    cm_alpha, cm_beta, cm_gamma, cm_delta,
    name = "SyntheticSet"
)
cms_syn
```

A list of CMs works equally well:

```{r construct-list, message = FALSE}
cm_list <- list(cm_alpha, cm_beta)
cms_small <- ConsortiumMetabolismSet(
    cm_list,
    name = "SmallSet"
)
cms_small
```

## From MiSoSoup data

The `misosoup24` dataset contains 56 metabolic solutions. Here we
construct CMs from the first four and combine them into a CMS.

```{r construct-misosoup, message = FALSE}
data("misosoup24")

## Build 4 CMs from the dataset
cm_ms <- lapply(seq_len(4), function(i) {
    ConsortiumMetabolism(
        misosoup24[[i]],
        name = names(misosoup24)[i],
        species_col = "species",
        metabolite_col = "metabolites",
        flux_col = "fluxes"
    )
})

cms_ms <- ConsortiumMetabolismSet(
    cm_ms,
    name = "MiSoSoup_subset"
)
cms_ms
```

# The `show()` method

Printing a CMS displays the name, the number of consortia, and the
description.

```{r show}
cms_syn
```

# Accessors

## `name()` and `description()`

```{r acc-name}
name(cms_syn)
description(cms_syn)
```

## `species()`

For a CMS, `species()` returns a tibble of all species across
consortia with the number of distinct pathways each species catalyses,
sorted in descending order.

```{r acc-species}
species(cms_syn)
```

## `metabolites()`

Returns a sorted character vector of all metabolites across the
universal metabolite space.

```{r acc-metabolites}
metabolites(cms_syn)
```

## `pathways()`

Returns pathway information across all consortia. The `type` argument
filters to specific subsets.

```{r acc-pathways-all}
## All pathways with consortium counts
pw <- pathways(cms_syn)
head(pw)
```

# Overlap matrix

The overlap matrix stores pairwise dissimilarity scores
(1 - Functional Overlap Score) between all consortia. The FOS is the
Szymkiewicz-Simpson coefficient applied to the binary
metabolite-metabolite matrices.

```{r overlap-matrix}
cms_syn@OverlapMatrix
```

Values near 0 on the diagonal indicate self-identity; off-diagonal
values close to 0 mean high similarity.

# Dendrogram

The dendrogram is built from the overlap (dissimilarity) matrix using
hierarchical clustering and is stored in the `Dendrogram` slot.

```{r dendrogram}
dend <- cms_syn@Dendrogram[[1]]
plot(dend, main = "Consortium dendrogram")
```

The `NodeData` slot contains the positions of internal dendrogram
nodes, which are used by `extractCluster()` and `plot()`.

```{r node-data}
cms_syn@NodeData
```

# Pathway classification

The `pathways()` method with a `type` argument classifies pathways by
prevalence across consortia or species.

## Pan-consortia pathways

Pathways present in many consortia (top fraction).

```{r pathways-pancons}
pathways(cms_syn, type = "pan-cons")
```

## Niche pathways

Pathways present in few consortia (bottom fraction).

```{r pathways-niche}
pathways(cms_syn, type = "niche")
```

## Core pathways

Pathways catalysed by many species (top fraction).

```{r pathways-core}
pathways(cms_syn, type = "core")
```

## Auxiliary pathways

Pathways catalysed by few species (bottom fraction).

```{r pathways-aux}
pathways(cms_syn, type = "aux")
```

The `quantileCutoff` parameter controls the threshold (default 0.1).

# Species classification

The `species()` method for CMS supports filtering by metabolic role.

## Generalists

Species with the most distinct pathways (top 15% by default).

```{r species-generalists}
species(cms_syn, type = "generalists")
```

## Specialists

Species with the fewest distinct pathways (bottom 15% by default).

```{r species-specialists}
species(cms_syn, type = "specialists")
```

The `quantileCutoff` parameter adjusts the fraction:

```{r species-cutoff}
species(cms_syn, type = "generalists", quantileCutoff = 0.25)
```

# Functional groups

`functionalGroups()` computes pairwise Jaccard similarities between
species based on their pathway sets, performs hierarchical clustering,
and returns a dendrogram coloured by `k` groups.

```{r functional-groups, fig.cap = "Functional groups dendrogram."}
fg <- functionalGroups(cms_syn, k = 3)
```

The returned list (invisible) contains the plot, dendrogram, similarity
matrix, species combinations, and per-species reaction sets:

```{r fg-contents}
names(fg)
```

# Cluster extraction

Given the numbered internal nodes visible in `plot(cms)` or
`NodeData`, `extractCluster()` pulls out a sub-CMS containing only the
consortia under a given node.

```{r extract-cluster, message = FALSE}
sub_cms <- extractCluster(cms_syn, node_id = 1)
sub_cms
```

# Replacement methods

## `name<-`

```{r replace-name}
name(cms_syn) <- "RenamedSet"
name(cms_syn)
```

## `description<-`

```{r replace-desc}
description(cms_syn) <- "Four synthetic consortia"
description(cms_syn)
```

# Session info

```{r session-info}
sessionInfo()
```
