---
title: "Alignment of Microbial Consortia"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Alignment of Microbial Consortia}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

# Introduction

The `ramen` package provides tools for comparing microbial communities based on
their metabolic exchange networks. The **alignment system** quantifies how
similar two or more communities are from a functional perspective -- not by
which species are present, but by which metabolite-metabolite pathways they
catalyse.

This vignette covers:

1. **Pairwise alignment** -- comparing two communities
2. **Multiple alignment** -- comparing all communities in a set
3. **Visualization** -- heatmaps, networks, and score plots
4. **Programmatic access** -- extracting results with accessor functions

```{r load-library, message = FALSE}
library(ramen)
```

# Creating test data

We use `synCM()` to generate synthetic communities with reproducible random
metabolic networks. Each consortium has species with random flux values
across a shared metabolite pool.

```{r create-data}
cm_alpha <- synCM("Alpha", n_species = 5, max_met = 10, seed = 42)
cm_beta <- synCM("Beta", n_species = 5, max_met = 10, seed = 43)
cm_gamma <- synCM("Gamma", n_species = 4, max_met = 8, seed = 44)
cm_delta <- synCM("Delta", n_species = 6, max_met = 10, seed = 45)

cm_alpha
```

# Pairwise alignment

## Basic usage

`align(CM, CM)` compares two `ConsortiumMetabolism` objects and returns a
`ConsortiumMetabolismAlignment` (CMA) object.

```{r pairwise-basic}
cma <- align(cm_alpha, cm_beta)
cma
```

## Similarity metrics

Five metrics are available via the `method` argument. All four individual
metrics are always computed; `method` selects which becomes the primary score.

```{r pairwise-metrics}
cma_fos <- align(cm_alpha, cm_beta, method = "FOS")
cma_jac <- align(cm_alpha, cm_beta, method = "jaccard")
cma_bc <- align(cm_alpha, cm_beta, method = "brayCurtis")
cma_ro <- align(cm_alpha, cm_beta, method = "redundancyOverlap")

## All scores are stored regardless of method choice
scores(cma_fos)
```

| Metric | What it measures | Formula |
|--------|-----------------|---------|
| FOS | Structural overlap (Szymkiewicz-Simpson) | \|X AND Y\| / min(\|X\|, \|Y\|) |
| Jaccard | Symmetric set similarity | \|X AND Y\| / \|X OR Y\| |
| Bray-Curtis | Flux-weighted similarity | 1 - sum\|x-y\| / sum(x+y) |
| Redundancy Overlap | Species labor distribution | weighted Jaccard on nEdges |
| MAAS | Composite (0.4 FOS + 0.2 each) | weighted mean |

## MAAS composite score

The Metabolic Alignment Aggregate Score combines all four metrics. Weights
are renormalized when some metrics are unavailable (e.g., unweighted networks).

```{r pairwise-maas}
cma_maas <- align(cm_alpha, cm_beta, method = "MAAS")
cma_maas@PrimaryScore
```

## Pathway correspondences

The alignment classifies every metabolite-metabolite edge as shared, unique
to the query, or unique to the reference.

```{r pairwise-pathways}
sp <- sharedPathways(cma)
head(sp[, c("consumed", "produced")])

up <- uniquePathways(cma)
nrow(sp) # shared
nrow(up$query) # unique to Alpha
nrow(up$reference) # unique to Beta
```

## Permutation p-values

Statistical significance is assessed by degree-preserving network rewiring.
The query network's edges are shuffled while preserving each metabolite's
degree, and the metric is recomputed under the null.

```{r pairwise-pvalue}
cma_p <- align(
  cm_alpha,
  cm_beta,
  method = "FOS",
  computePvalue = TRUE,
  nPermutations = 99L
)
cma_p@Pvalue
```

# Multiple alignment

## Aligning a consortium set

`align(CMS)` computes pairwise similarities across all consortia in a
`ConsortiumMetabolismSet` and returns a CMA with `Type = "multiple"`.

```{r multiple-create, message = FALSE}
cms <- ConsortiumMetabolismSet(
  list(cm_alpha, cm_beta, cm_gamma, cm_delta),
  name = "Demo"
)
cma_mult <- align(cms)
cma_mult
```

## Similarity matrix

The `SimilarityMatrix` is an n x n symmetric matrix with 1s on the diagonal.
For FOS, this is derived directly from the pre-computed CMS overlap matrix,
guaranteeing numerical consistency.

```{r multiple-simmat}
round(similarityMatrix(cma_mult), 3)
```

## Summary scores

The `PrimaryScore` for a multiple alignment is the **median** of all
pairwise scores. Full summary statistics are available via `scores()`.

```{r multiple-scores}
scores(cma_mult)
```

## Consensus network and prevalence

Edge prevalence counts how many consortia share each metabolite-metabolite
pathway. This enables classification of edges as core (present in most
consortia) or niche (present in few).

```{r multiple-prevalence}
prev <- prevalence(cma_mult)
head(prev[order(-prev$nConsortia), ])

## Distribution
table(prev$nConsortia)
```

# Visualization

## Heatmap (multiple alignment)

The heatmap shows pairwise similarities with dendrogram-based ordering.

```{r plot-heatmap, fig.cap = "Similarity heatmap across four consortia."}
plot(cma_mult, type = "heatmap")
```

## Network (pairwise alignment)

The network view shows shared (green), query-unique (blue), and
reference-unique (red) pathways as a directed metabolite flow graph.

```{r plot-network, fig.cap = "Pathway network: Alpha vs Beta."}
plot(cma, type = "network")
```

## Score comparison

The scores bar chart works for both pairwise and multiple alignments.

```{r plot-scores-pairwise, fig.cap = "Pairwise metric scores."}
plot(cma, type = "scores")
```

```{r plot-scores-multiple, fig.cap = "Multiple alignment summary scores."}
plot(cma_mult, type = "scores")
```

# Programmatic access

All CMA results are accessible via noun-style accessors with type guards
that prevent misuse (e.g., calling `sharedPathways()` on a multiple
alignment raises an informative error).

| Accessor | Type | Returns |
|----------|------|---------|
| `scores()` | both | Named list of scores |
| `sharedPathways()` | pairwise | data.frame of shared edges |
| `uniquePathways()` | pairwise | list(query, reference) |
| `similarityMatrix()` | multiple | n x n numeric matrix |
| `prevalence()` | multiple | data.frame with nConsortia |
| `consensusEdges()` | multiple | data.frame (same as prevalence) |
| `metabolites()` | both | Character vector of metabolites |
| `pathways()` | both | data.frame of edges |

# Session info

```{r session-info}
sessionInfo()
```
