---
title: "The ConsortiumMetabolism Class"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{The ConsortiumMetabolism Class}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

# Introduction

The `ConsortiumMetabolism` (CM) class is the fundamental building block
of the `ramen` package. It represents a single microbial consortium's
metabolic exchange network -- the directed graph of metabolite
consumption and production by its member species.

A CM object stores:

- **Six assay matrices** (square, metabolite-by-metabolite, sparse):
  Binary, nSpecies, Consumption, Production, EffectiveConsumption,
  EffectiveProduction
- **Pathway data** linking consumed and produced metabolites with
  per-pathway summary statistics
- **An igraph object** of the directed metabolic network

This vignette covers construction, accessors, replacement methods,
subsetting, and inspection of CM objects.

```{r load-library, message = FALSE}
library(ramen)
```

# Construction

## From an edge list

The `ConsortiumMetabolism()` constructor accepts a data.frame (or
tibble) with columns for species, metabolite, and flux. Negative fluxes
indicate consumption; positive fluxes indicate production.

```{r construct-from-df}
edge_data <- data.frame(
    species = c(
        "Sp_A", "Sp_A", "Sp_A", "Sp_A",
        "Sp_B", "Sp_B", "Sp_B",
        "Sp_C", "Sp_C", "Sp_C"
    ),
    met = c(
        "met1", "met2", "met3", "met4",
        "met1", "met3", "met4",
        "met2", "met3", "met4"
    ),
    flux = c(
        -0.30, 0.53, -2.23, 3.31,
        2.55, 0.34, -1.85,
        -1.30, -0.48, 0.60
    )
)

cm_manual <- ConsortiumMetabolism(
    edge_data,
    name = "manual_example"
)
cm_manual
```

Custom column names can be specified via `species_col`, `metabolite_col`,
and `flux_col`:

```{r construct-custom-cols}
df <- data.frame(
    organism = c("Sp_A", "Sp_A", "Sp_B", "Sp_B"),
    compound = c("glc", "ac", "glc", "ac"),
    rate = c(-1.0, 0.5, 0.8, -0.3)
)

cm_custom <- ConsortiumMetabolism(
    df,
    name = "custom_cols",
    species_col = "organism",
    metabolite_col = "compound",
    flux_col = "rate"
)
cm_custom
```

## With `synCM()`

For quick testing and demonstration, `synCM()` generates synthetic
consortia with random species-metabolite interactions.

```{r construct-synCM}
cm_syn <- synCM(
    "SyntheticComm",
    n_species = 5,
    max_met = 8,
    seed = 42
)
cm_syn
```

To retrieve the raw edge list without constructing a CM object,
set `cm = FALSE`:

```{r synCM-raw}
raw_edges <- synCM(
    "raw",
    n_species = 3,
    max_met = 5,
    seed = 1,
    cm = FALSE
)
head(raw_edges)
```

## From MiSoSoup data

The package ships with `misosoup24`, a list of 56 metabolic solutions
from [MiSoSoup](https://github.com/sirno/misosoup). Each element is a
tibble with columns `metabolites`, `species`, and `fluxes`.

```{r construct-misosoup}
data("misosoup24")
length(misosoup24)
names(misosoup24)[1:6]

## Create a CM from the first solution
cm_ms <- ConsortiumMetabolism(
    misosoup24[[1]],
    name = names(misosoup24)[1],
    species_col = "species",
    metabolite_col = "metabolites",
    flux_col = "fluxes"
)
cm_ms
```

# The `show()` method

Printing a CM object gives a concise summary including the name,
whether the network is weighted, and the number of metabolites.

```{r show}
cm_syn
```

# Accessors

## `name()` and `description()`

```{r acc-name}
name(cm_syn)
description(cm_syn)
```

## `species()`

Returns the unique species identifiers.

```{r acc-species}
species(cm_syn)
```

## `metabolites()`

Returns the unique metabolite identifiers.

```{r acc-metabolites}
metabolites(cm_syn)
```

## `pathways()`

Returns the pathway data.frame, which contains per-pathway summary
statistics including the number of species catalysing each pathway,
summed consumption and production fluxes, and effective diversity.

```{r acc-pathways}
pw <- pathways(cm_syn)
pw[, c("consumed", "produced", "n_species")]
```

## `consortia()`
Returns the input data in a tidy format with columns `met`, `species`,
and `flux`.

```{r acc-consortia}
head(consortia(cm_syn))
```

# Replacement methods

## `name<-`

```{r replace-name}
name(cm_syn) <- "RenamedComm"
name(cm_syn)
```

## `description<-`

```{r replace-desc}
description(cm_syn) <- "A synthetic test community"
description(cm_syn)
```

# Assay matrices

A CM object inherits from `TreeSummarizedExperiment` and stores six
square (metabolite-by-metabolite) sparse assay matrices. Rows represent
consumed metabolites and columns represent produced metabolites.

```{r assay-names}
SummarizedExperiment::assayNames(cm_ms)
```

Each matrix encodes a different view of the metabolic network:

| Assay | Content |
|-------|---------|
| Binary | 1 if a pathway exists, 0 otherwise |
| nSpecies | Number of species catalysing the pathway |
| Consumption | Summed consumption flux across species |
| Production | Summed production flux across species |
| EffectiveConsumption | Effective number of consuming species |
| EffectiveProduction | Effective number of producing species |

```{r assay-binary}
## The Binary assay (sparse matrix)
SummarizedExperiment::assay(cm_ms, "Binary")
```

# Dimensionality

Because CM inherits from `TreeSummarizedExperiment`, standard dimension
accessors work:

```{r dim}
dim(cm_ms)
nrow(cm_ms)
ncol(cm_ms)
```

The rows and columns both correspond to the metabolite universe of
this consortium (including a synthetic "media" node when needed).

# Session info

```{r session-info}
sessionInfo()
```
