---
title: "Introduction to ramen"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Introduction to ramen}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 5
)
```

# Installation

`ramen` can be installed from Bioconductor:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ramen")
```

```{r load-library, message = FALSE}
library(ramen)
```

# Overview

`ramen` (**R**econstruction and **A**lignment of **M**icrobial
**E**xchange **N**etworks) analyses the functional similarity of
microbial communities from a metabolic network perspective. Rather
than comparing communities by species composition, `ramen` asks:
*which metabolite-to-metabolite pathways does each community
catalyse, and how similar are the resulting networks?*

The package provides three S4 classes that build on Bioconductor's
`TreeSummarizedExperiment`:
`r BiocStyle::Biocpkg("TreeSummarizedExperiment")`:

- **`ConsortiumMetabolism`** (CM) -- a single community's metabolic
  exchange network
- **`ConsortiumMetabolismSet`** (CMS) -- a collection of CMs with
  precomputed overlap scores and a dendrogram
- **`ConsortiumMetabolismAlignment`** (CMA) -- the result of
  aligning two or more communities

This vignette walks through a complete analysis from data import to
alignment using the bundled `misosoup24` dataset (56 metabolic
solutions from [MiSoSoup](https://github.com/sirno/misosoup)). For
details on alignment metrics, see
`vignette("alignment", package = "ramen")`; for a gallery of all
plot types, see `vignette("visualisation", package = "ramen")`.

# Data import

## The `misosoup24` dataset

`ramen` ships with `misosoup24`, a list of 56 metabolic solutions.
Each element is a data.frame with columns `metabolites`, `species`,
and `fluxes`, where negative fluxes indicate consumption and
positive fluxes indicate production.

```{r load-data}
data("misosoup24")
length(misosoup24)
names(misosoup24)[1:8]
head(misosoup24[[1]])
```

## Importing raw MiSoSoup YAML

For raw MiSoSoup output (nested YAML), `importMisosoup()` parses
the data into structured tibbles of consortia, media, and growth
information. `overviewMisosoup()` provides a quick summary of the
data structure before full import.

```{r import-misosoup, eval = FALSE}
## Parse raw MiSoSoup YAML (not run -- requires external data)
raw <- yaml::read_yaml("path/to/misosoup_output.yaml")
overviewMisosoup(raw)
result <- importMisosoup(raw)
names(result) # "consortia", "media", "growth"
```

## Alternative input formats

The `ConsortiumMetabolism()` constructor accepts any data.frame with
columns for species, metabolite, and flux. Custom column names are
specified via `species_col`, `metabolite_col`, and `flux_col`.

If your data is in wide format with separate columns for consumed
and produced metabolites, `pivotCM()` converts it to the long
format that `ConsortiumMetabolism()` expects:

```{r pivot-cm}
wide_data <- data.frame(
    species = c("Sp_A", "Sp_B", "Sp_C"),
    uptake = c("met1", "met2", "met3"),
    secretion = c("met2", "met3", "met1"),
    flux = c(1, 1, 1)
)

long_data <- pivotCM(
    wide_data,
    species = "species",
    from = "uptake",
    to = "secretion",
    flux = "flux"
)
head(long_data)
```

# Building a ConsortiumMetabolism

## From `misosoup24`

```{r build-cm}
cm1 <- ConsortiumMetabolism(
    misosoup24[[1]],
    name = names(misosoup24)[1],
    species_col = "species",
    metabolite_col = "metabolites",
    flux_col = "fluxes"
)
cm1
```

## Synthetic data with `synCM()`

For quick testing, `synCM()` generates a random consortium:

```{r syn-cm}
cm_syn <- synCM("Synthetic", n_species = 4, max_met = 8, seed = 42)
cm_syn
```

## Inspecting a CM

A CM inherits from `TreeSummarizedExperiment`, so standard
dimension accessors work. Rows and columns both correspond to
metabolites (it is a square metabolite-by-metabolite matrix):

```{r cm-dim}
dim(cm1)
```

The six assay matrices encode different views of the network:

```{r cm-assays}
SummarizedExperiment::assayNames(cm1)
```

| Assay | Content |
|-------|---------|
| Binary | 1 if a pathway exists, 0 otherwise |
| nSpecies | Number of species catalysing the pathway |
| Consumption | Summed consumption flux |
| Production | Summed production flux |
| EffectiveConsumption | Effective number of consuming species |
| EffectiveProduction | Effective number of producing species |

## Accessors

```{r cm-accessors}
name(cm1)
species(cm1)
metabolites(cm1)
```

`pathways()` returns per-pathway summary statistics:

```{r cm-pathways}
head(pathways(cm1))
```

`consortia()` returns the tidy input data:

```{r cm-consortia}
head(consortia(cm1))
```

Replacement methods allow renaming:

```{r cm-replace}
name(cm1) <- "first_solution"
name(cm1)
```

# Building a ConsortiumMetabolismSet

A `ConsortiumMetabolismSet` groups multiple CMs. During
construction, `ramen` automatically expands all binary matrices to a
shared universal metabolite space, computes pairwise overlap scores,
and builds a hierarchical clustering dendrogram.

```{r build-cms, message = FALSE}
## Build 20 CMs from misosoup24
cm_list <- lapply(seq_len(20), function(i) {
    ConsortiumMetabolism(
        misosoup24[[i]],
        name = names(misosoup24)[i],
        species_col = "species",
        metabolite_col = "metabolites",
        flux_col = "fluxes"
    )
})

cms <- ConsortiumMetabolismSet(cm_list, name = "MiSoSoup_20")
cms
```

## Species across the set

`species()` returns all species with the number of distinct pathways
each catalyses:

```{r cms-species}
species(cms)
```

Filter by metabolic role -- generalists (top fraction by pathway
count) or specialists (bottom fraction):

```{r cms-species-types}
species(cms, type = "generalists")
species(cms, type = "specialists")
```

## Pathway classification

`pathways()` with a `type` argument classifies pathways by
prevalence:

```{r cms-pathways-types}
## Pan-consortia: present in many consortia (top fraction)
head(pathways(cms, type = "pan-cons"))

## Niche: present in few consortia (bottom fraction)
head(pathways(cms, type = "niche"))
```

The `quantileCutoff` parameter controls the threshold (default
0.1).

## Functional groups

`functionalGroups()` clusters species by the Jaccard similarity of
their pathway sets:

```{r cms-func-groups, fig.cap = "Functional groups dendrogram."}
fg <- functionalGroups(cms, k = 3)
```

## Dendrogram and cluster extraction

```{r cms-dendrogram, fig.cap = "CMS dendrogram with numbered nodes."}
plot(cms)
```

The numbered internal nodes can be used to extract sub-clusters:

```{r cms-extract, message = FALSE}
sub_cms <- extractCluster(cms, node_id = 1)
sub_cms
```

# Alignment

## Pairwise alignment

`align()` compares two CMs and returns a
`ConsortiumMetabolismAlignment`:

```{r align-pairwise}
cma_pair <- align(cm_list[[1]], cm_list[[2]])
cma_pair
```

All similarity metrics are computed automatically:

```{r align-scores}
scores(cma_pair)
```

Pathway correspondences show shared and unique pathways:

```{r align-pathways}
shared <- pathways(cma_pair, type = "shared")
unique_pw <- pathways(cma_pair, type = "unique")
nrow(shared)
nrow(unique_pw$query)
nrow(unique_pw$reference)
```

## Multiple alignment

`align()` on a CMS computes all pairwise similarities:

```{r align-multiple, message = FALSE}
cma_mult <- align(cms)
cma_mult
```

The similarity matrix and summary scores:

```{r align-mult-results}
scores(cma_mult)
round(similarityMatrix(cma_mult)[1:5, 1:5], 3)
```

Pathway prevalence across consortia:

```{r align-prevalence}
prev <- prevalence(cma_mult)
head(prev[order(-prev$nConsortia), ])
```

For a detailed treatment of alignment metrics, p-values, and
accessor functions, see
`vignette("alignment", package = "ramen")`.

# Visualization

Each class has a `plot()` method. Here is one example per class;
for the full gallery, see
`vignette("visualisation", package = "ramen")`.

```{r plot-cm, fig.cap = "Metabolic network for the first consortium."}
plot(cm_list[[1]], type = "Binary")
```

```{r plot-cms, fig.cap = "Dendrogram of 20 consortia."}
plot(cms)
```

```{r plot-cma, fig.cap = "Pairwise similarity heatmap."}
plot(cma_mult, type = "heatmap")
```

# Session info

```{r session-info}
sessionInfo()
```
