---
title: "Introduction to ramen"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Introduction to ramen}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 5
)
```

# Installation

`ramen` can be installed from Bioconductor:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ramen")
```

```{r load-library, message = FALSE}
library(ramen)
```

# Overview

`ramen` (**R**econstruction and **A**lignment of **M**icrobial
**E**xchange **N**etworks) analyses the functional similarity of
microbial communities from a metabolic network perspective. Rather
than comparing communities by species composition, `ramen` asks:
*which metabolite-to-metabolite pathways does each community
catalyse, and how similar are the resulting networks?*

## Key concept: pathways

Throughout `ramen`, a **pathway** is a directed edge from one
metabolite to another: "metabolite A is consumed and metabolite B
is produced by at least one species." This is *not* a biochemical
pathway in the KEGG or MetaCyc sense (e.g. glycolysis). Rather, it
captures a single metabolic exchange coupling within the community.
A consortium's full set of pathways forms a square
metabolite-by-metabolite network that encodes its collective
metabolic capability.

## Package classes

The package provides three S4 classes that build on Bioconductor's
`r BiocStyle::Biocpkg("TreeSummarizedExperiment")`:

- **`ConsortiumMetabolism`** (CM) -- a single community's metabolic
    exchange network
- **`ConsortiumMetabolismSet`** (CMS) -- a collection of CMs with
    precomputed overlap scores and a dendrogram
- **`ConsortiumMetabolismAlignment`** (CMA) -- the result of
    aligning two or more communities

This vignette walks through a complete analysis from data import to
alignment using the bundled `misosoup24` dataset (56 metabolic
solutions from [MiSoSoup](https://github.com/sirno/misosoup)). For
details on alignment metrics, see
`vignette("alignment", package = "ramen")`; for a gallery of all
plot types, see `vignette("visualisation", package = "ramen")`.

## Why "consortium" and not "community"?

`ramen` deliberately uses the term *consortium* throughout. A
consortium here is defined by its **metabolic function** -- the set
of metabolite-to-metabolite pathways its member species
collectively catalyse -- not by its taxonomic composition. Two
assemblages with entirely different species rosters but identical
exchange networks are, from `ramen`'s perspective, the same
consortium; conversely, two assemblages with the same species but
different active pathways are different consortia. This functional
framing follows a long-standing argument in microbial ecology that
ecosystem behaviour is more robustly predicted by what microbes
*do* than by which microbes are present (Falkowski, Fenchel &
Delong 2008, *Science* **320**:1034).

# Key Concepts

A short glossary of terms used throughout the package:

- **Consortium** -- a microbial assemblage characterised by its
    metabolic exchange network, not its species composition (see
    above).
- **Pathway** -- in the `ramen` sense, a directed
    metabolite-to-metabolite edge: metabolite *A* is consumed and
    metabolite *B* is produced by at least one species. Not a
    KEGG/MetaCyc biochemical pathway.
- **Assay** -- one of the six *m x m* metabolite-by-metabolite
    matrices stored in a `ConsortiumMetabolism` object (Binary,
    nSpecies, Consumption, Production, EffectiveConsumption,
    EffectiveProduction).
- **FOS** -- Functional Overlap Score; the primary similarity
    metric between two consortia, computed as the
    Szymkiewicz-Simpson coefficient on expanded binary pathway
    matrices.
- **Szymkiewicz-Simpson** -- an asymmetric set-overlap coefficient
    `|X intersect Y| / min(|X|, |Y|)` that measures how completely
    the smaller set is contained in the larger.
- **Alignment** -- the operation of putting two or more consortia
    in correspondence in a shared metabolite/pathway space and
    summarising their similarity, returning a
    `ConsortiumMetabolismAlignment`.
- **MAAS** -- Metabolite Abundance Adjusted Score; a
    flux-magnitude-weighted variant of FOS that down-weights
    pathways with very small fluxes.

# Data import

## The `misosoup24` dataset

`ramen` ships with `misosoup24`, a list of 56 metabolic solutions.
Each element is a data.frame with columns `metabolites`, `species`,
and `fluxes`, where negative fluxes indicate consumption and
positive fluxes indicate production.

> **Metabolite identifiers** follow the
> [BiGG namespace](https://bigg.ucsd.edu) (King et al. 2016):
> short lowercase codes such as `ac` (acetate), `co2` (CO~2~),
> `etoh` (ethanol), and `pyr` (pyruvate). Double-underscores
> encode stereochemistry: `ala__L` = L-alanine, `ala__D` =
> D-alanine. Look up any identifier at
> <https://bigg.ucsd.edu/metabolites>.

```{r load-data}
data("misosoup24")
length(misosoup24)
names(misosoup24)[1:8]
head(misosoup24[[1]])
```

> **Caution -- alternative optima.** MiSoSoup is a Mixed Integer
> Linear Programming (MILP) enumerator: for a single underlying
> metabolic model it returns *multiple alternative optimal
> solutions* to the same growth problem. Two such alternatives will
> typically share most of their pathways and produce very high
> pairwise FOS values. This high overlap is **expected by
> construction** and reflects solver consistency, not biological
> similarity. When interpreting overlap scores, compare consortia
> derived from *different* models or conditions; treat overlap
> between alternatives of the same model as a sanity check on the
> enumeration, not as an ecological signal.

## Importing raw MiSoSoup YAML

For raw MiSoSoup output (nested YAML), `importMisosoup()` parses the
data directly into a `ConsortiumMetabolismSet` containing one
`ConsortiumMetabolism` per viable solution. The function accepts a
single YAML file, a directory of YAML files, or a pre-loaded nested
list from `yaml::read_yaml()`. Media-level exchange bounds are stashed
in each CM's `metadata()` slot under `$media` so they remain
accessible if needed.

```{r import-misosoup, eval = FALSE}
## Single file -> CMS (not run -- requires external data)
cms <- importMisosoup("path/to/misosoup_output.yaml")

## Directory of YAML files -> CMS merging all consortia
cms <- importMisosoup("path/to/misosoup_dir/", name = "experiment1")

## Pre-loaded list -> CMS (name required)
raw <- yaml::read_yaml("path/to/misosoup_output.yaml")
cms <- importMisosoup(raw, name = "experiment1")

## Access stashed media bounds for a specific CM
## S4Vectors::metadata(cms@Consortia[[1]])$media
```

## Alternative input formats

The `ConsortiumMetabolism()` constructor accepts any data.frame with
columns `species`, `metabolite`, and `flux`. Custom column names can
be specified via `species_col`, `metabolite_col`, and `flux_col`.

If your data is in wide format with separate columns for consumed
and produced metabolites, `pivotCM()` converts it to the long
format that `ConsortiumMetabolism()` expects:

```{r pivot-cm}
wide_data <- data.frame(
    species = c("Sp_A", "Sp_B", "Sp_C"),
    uptake = c("met1", "met2", "met3"),
    secretion = c("met2", "met3", "met1"),
    flux = c(1, 1, 1)
)

long_data <- pivotCM(
    wide_data,
    species = "species",
    from = "uptake",
    to = "secretion",
    flux = "flux"
)
head(long_data)
```

## Importing from MICOM / cobrapy

`ConsortiumMetabolism()` accepts any long-format `data.frame` with
columns `species`, `metabolite`, and `flux` (negative = consumption,
positive = production). This makes interoperation with the
Python-based [MICOM](https://github.com/micom-dev/micom) and
[cobrapy](https://github.com/opencobra/cobrapy) ecosystems
straightforward: export per-species exchange fluxes to CSV from
Python, then read into R.

A minimal MICOM growth solution can be exported as follows:

```{python micom-export, eval = FALSE, python.reticulate = FALSE}
import pandas as pd
from micom import Community
from micom.workflows import grow

## com is a pre-built micom.Community
sol = com.cooperative_tradeoff()

## sol.exchanges is a tidy DataFrame with columns
##   reaction, metabolite, taxon (species), flux, ...
exchanges = sol.exchanges[["taxon", "metabolite", "flux"]]
exchanges.columns = ["species", "metabolite", "flux"]
exchanges.to_csv("micom_exchanges.csv", index = False)
```

The same pattern works for a plain `cobra.Model`: iterate over
`model.exchanges`, record the carrying species, and write the
columns above.

On the R side:

```{r micom-import, eval = FALSE}
exch <- read.csv("micom_exchanges.csv")
cm_micom <- ConsortiumMetabolism(exch, name = "micom_run1")
```

# Building a ConsortiumMetabolism

## From `misosoup24`

```{r build-cm}
cm1 <- ConsortiumMetabolism(
    misosoup24[[1]],
    name = names(misosoup24)[1]
)
cm1
```

## Synthetic data with `synCM()`

For quick testing, `synCM()` generates a random consortium:

```{r syn-cm}
cm_syn <- synCM("Synthetic", n_species = 4, max_met = 8, seed = 42)
cm_syn
```

## Inspecting a CM

A CM inherits from `TreeSummarizedExperiment`, so standard
dimension accessors work. Rows and columns both correspond to
metabolites (it is a square metabolite-by-metabolite matrix):

```{r cm-dim}
dim(cm1)
```

The six assay matrices encode different views of the network:

```{r cm-assays}
SummarizedExperiment::assayNames(cm1)
```

| Assay | Content |
|-------|---------|
| Binary | 1 if a pathway exists, 0 otherwise |
| nSpecies | Number of species catalysing the pathway |
| Consumption | Summed consumption flux |
| Production | Summed production flux |
| EffectiveConsumption | Effective number of consuming species |
| EffectiveProduction | Effective number of producing species |

## Accessors

```{r cm-accessors}
name(cm1)
species(cm1)
metabolites(cm1)
```

`pathways()` returns per-pathway summary statistics:

```{r cm-pathways}
head(pathways(cm1))
```

`as.data.frame()` returns the underlying edge list. Use this to
export a CM's data for downstream tools:

```{r cm-as-data-frame}
head(as.data.frame(cm1))
```

(`consortia()` is reserved for containers of CMs --
`ConsortiumMetabolismSet` and `ConsortiumMetabolismAlignment` --
since the noun is plural.)

Replacement methods allow renaming:

```{r cm-replace}
name(cm1) <- "first_solution"
name(cm1)
```

# Building a ConsortiumMetabolismSet

A `ConsortiumMetabolismSet` groups multiple CMs. During
construction, `ramen` automatically expands all binary matrices to a
shared universal metabolite space, computes pairwise overlap scores,
and builds a hierarchical clustering dendrogram.

```{r build-cms, message = FALSE}
## Build 20 CMs from misosoup24
cm_list <- lapply(seq_len(20), function(i) {
    ConsortiumMetabolism(
        misosoup24[[i]],
        name = names(misosoup24)[i]
    )
})

cms <- ConsortiumMetabolismSet(cm_list, name = "MiSoSoup_20")
cms
```

## Species across the set

`species()` returns all species with the number of distinct pathways
each catalyses:

```{r cms-species}
species(cms)
```

Filter by metabolic role -- generalists (top fraction by pathway
count) or specialists (bottom fraction):

```{r cms-species-types}
species(cms, type = "generalists")
species(cms, type = "specialists")
```

## Pathway classification

`pathways()` with a `type` argument classifies pathways by
prevalence:

```{r cms-pathways-types}
## Pan-consortia: present in many consortia (top fraction)
head(pathways(cms, type = "pan-cons"))

## Niche: present in few consortia (bottom fraction)
head(pathways(cms, type = "niche"))
```

The `quantileCutoff` parameter controls the threshold (default
0.1).

## Functional groups

`functionalGroups()` clusters species by the Jaccard similarity of
their pathway sets. `plotFunctionalGroups()` visualizes the
resulting dendrogram:

```{r cms-func-groups, fig.cap = "Functional groups dendrogram."}
fg <- functionalGroups(cms)
plotFunctionalGroups(fg, k = 3)
```

## Dendrogram and cluster extraction

```{r cms-dendrogram, fig.cap = "CMS dendrogram with numbered nodes."}
plot(cms)
```

The numbered internal nodes can be used to extract sub-clusters:

```{r cms-extract, message = FALSE}
sub_cms <- extractCluster(cms, node_id = 1)
sub_cms
```

# Alignment

## Pairwise alignment

`align()` compares two CMs and returns a
`ConsortiumMetabolismAlignment`:

```{r align-pairwise}
cma_pair <- align(cm_list[[1]], cm_list[[2]])
cma_pair
```

All similarity metrics are computed automatically:

```{r align-scores}
scores(cma_pair)
```

Pathway correspondences show shared and unique pathways:

```{r align-pathways}
shared <- pathways(cma_pair, type = "shared")
unique_pw <- pathways(cma_pair, type = "unique")
nrow(shared)
nrow(unique_pw$query)
nrow(unique_pw$reference)
```

## Multiple alignment

`align()` on a CMS computes all pairwise similarities:

```{r align-multiple, message = FALSE}
cma_mult <- align(cms)
cma_mult
```

The similarity matrix and summary scores:

```{r align-mult-results}
scores(cma_mult)
round(similarityMatrix(cma_mult)[1:5, 1:5], 3)
```

Pathway prevalence across consortia:

```{r align-prevalence}
prev <- prevalence(cma_mult)
head(prev[order(-prev$nConsortia), ])
```

For a detailed treatment of alignment metrics, p-values, and
accessor functions, see
`vignette("alignment", package = "ramen")`.

# Visualization

Each class has a `plot()` method. Here is one example per class;
for the full gallery, see
`vignette("visualisation", package = "ramen")`.

```{r plot-cm, fig.cap = "Metabolic network for the first consortium."}
plot(cm_list[[1]], type = "Binary")
```

```{r plot-cms, fig.cap = "Dendrogram of 20 consortia."}
plot(cms)
```

```{r plot-cma, fig.cap = "Pairwise similarity heatmap."}
plot(cma_mult, type = "heatmap")
```

# Bioconductor / SummarizedExperiment interoperability

`ramen` classes inherit from
`r BiocStyle::Biocpkg("TreeSummarizedExperiment")`, so the standard
`r BiocStyle::Biocpkg("SummarizedExperiment")` accessors work out
of the box. `ramen` is a first-class Bioconductor citizen, not an
isolated framework -- you can mix its objects freely with other
SE-based tools.

Pull a raw assay matrix:

```{r tse-assay}
SummarizedExperiment::assay(cm1, "Binary")[1:5, 1:5]
```

Both rows and columns of a CM are metabolites (it is a square
*m x m* network), so `dim()`, `rowData()`, and `colData()` all act
in metabolite space:

```{r tse-cm-dim}
dim(cm1)
head(SummarizedExperiment::rowData(cm1))
head(SummarizedExperiment::colData(cm1))
```

A `ConsortiumMetabolismSet` likewise exposes the SE interface, with
columns indexing the universal metabolite space:

```{r tse-cms-dim}
dim(cms)
head(colnames(cms))
```

Subsetting works in metabolite space using the standard `[i, j]`
syntax -- here, the first 5 rows and 5 columns:

```{r tse-cms-subset}
cms[1:5, 1:5]
```

This means downstream Bioconductor tooling -- e.g. `mia` for
microbiome analysis or any package consuming `(Tree)SummarizedExperiment`
objects -- can operate on `ramen` outputs without any conversion.

# Session info

```{r session-info}
sessionInfo()
```
