---
title: "Visualisation"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Visualisation}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 5
)
```

# Introduction

`ramen` provides `plot()` methods for all three core classes:
`ConsortiumMetabolism`, `ConsortiumMetabolismSet`, and
`ConsortiumMetabolismAlignment`. This vignette is a visual gallery
of every plot type. For context on the underlying analyses, see
`vignette("ramen", package = "ramen")` and
`vignette("alignment", package = "ramen")`.

```{r load-library, message = FALSE}
library(ramen)
```

# Example data

```{r create-data, message = FALSE}
data("misosoup24")
cm_list <- lapply(seq_len(6), function(i) {
    ConsortiumMetabolism(
        misosoup24[[i]],
        name = names(misosoup24)[i]
    )
})

cms <- ConsortiumMetabolismSet(cm_list, name = "Demo")
cma_pair <- align(cm_list[[1]], cm_list[[2]])
cma_mult <- align(cms)
```

# ConsortiumMetabolism plots

`plot(CM)` renders a directed metabolic flow network using
`igraph`. Nodes are coloured by role: **lightblue** for sources
(only outgoing edges), **salmon** for sinks (only incoming edges),
and **lightgoldenrod** for intermediate nodes (both incoming and
outgoing). The `type` argument selects which assay matrix
determines edge weights.

## Binary network

The default view shows network structure with uniform edge weights.

```{r plot-cm-binary, fig.cap = "Binary metabolic network."}
plot(cm_list[[1]], type = "Binary")
```

## Number of species per pathway

Edge weights reflect how many species catalyse each pathway.

```{r plot-cm-nspecies, fig.cap = "nSpecies-weighted network."}
plot(cm_list[[1]], type = "nSpecies")
```

## Consumption and production

Edges weighted by summed flux magnitudes.

```{r plot-cm-consumption, fig.cap = "Consumption-weighted network."}
plot(cm_list[[1]], type = "Consumption")
```

```{r plot-cm-production, fig.cap = "Production-weighted network."}
plot(cm_list[[1]], type = "Production")
```

## Effective consumption and production

Effective diversity measures how evenly species contribute to each
pathway (Shannon-based effective number of species).

```{r plot-cm-effcons, fig.cap = "Effective consumption network."}
plot(cm_list[[1]], type = "EffectiveConsumption")
```

```{r plot-cm-effprod, fig.cap = "Effective production network."}
plot(cm_list[[1]], type = "EffectiveProduction")
```

# ConsortiumMetabolismSet plots

## Dendrogram

`plot(CMS)` draws a hierarchical clustering dendrogram with
numbered internal nodes. These node IDs are used by
`extractCluster()` to pull sub-clusters.

```{r plot-cms-dend, fig.cap = "CMS dendrogram with numbered nodes."}
plot(cms)
```

## Custom label colours

Supply a data.frame mapping leaf labels to colours.

```{r plot-cms-colours, fig.cap = "CMS dendrogram with coloured labels."}
colour_map <- data.frame(
    label = names(misosoup24)[seq_len(6)],
    colour = c(
        "steelblue", "firebrick", "forestgreen",
        "darkorange", "purple", "darkred"
    )
)
plot(cms, label_colours = colour_map)
```

# ConsortiumMetabolismAlignment plots

The CMA `plot()` method supports three types: `"heatmap"`,
`"network"`, and `"scores"`. The default is `"network"` for
pairwise and `"heatmap"` for multiple alignments.

## Heatmap (multiple alignment)

Pairwise FOS similarities with dendrogram-based ordering. Values
range from 0 (no overlap) to 1 (identical).

```{r plot-cma-heatmap, fig.cap = "Similarity heatmap."}
plot(cma_mult, type = "heatmap")
```

## Network (pairwise alignment)

Shared pathways in **green**, query-unique in **blue**,
reference-unique in **red**.

```{r plot-cma-network, fig.cap = "Pairwise alignment network."}
plot(cma_pair, type = "network")
```

## Scores

The bar chart displays similarity metrics for pairwise alignments
or summary statistics for multiple alignments.

```{r plot-cma-scores-pair, fig.cap = "Pairwise alignment scores."}
plot(cma_pair, type = "scores")
```

```{r plot-cma-scores-mult, fig.cap = "Multiple alignment summary scores."}
plot(cma_mult, type = "scores")
```

# Using `plotDirectedFlow()` directly

For fine-grained control over the network layout, call
`plotDirectedFlow()` directly on an igraph object. This is the
function underlying all CM and CMA network plots.

```{r plot-directed-flow, fig.cap = "Custom directed flow plot."}
g <- igraph::graph_from_adjacency_matrix(
    SummarizedExperiment::assay(cm_list[[2]], "nSpecies"),
    mode = "directed",
    weighted = TRUE
)

plotDirectedFlow(
    g,
    color_edges_by_weight = TRUE,
    edge_width_range = c(0.5, 2),
    vertex_size = 12,
    vertex_label_cex = 0.7,
    main = paste0(name(cm_list[[2]]), " - nSpecies (custom)")
)
```

# Plot type summary

| Object class | `type` argument | Description |
|-------------|-----------------|-------------|
| CM | `"Binary"` (default) | Directed network, unweighted |
| CM | `"nSpecies"` | Pathways weighted by species count |
| CM | `"Consumption"` | Pathways weighted by consumption flux |
| CM | `"Production"` | Pathways weighted by production flux |
| CM | `"EffectiveConsumption"` | Effective consuming species |
| CM | `"EffectiveProduction"` | Effective producing species |
| CMS | (none) | Dendrogram with numbered nodes |
| CMA | `"heatmap"` | Pairwise similarity heatmap |
| CMA | `"network"` | Shared/unique pathway graph |
| CMA | `"scores"` | Bar chart of similarity metrics |

# Session info

```{r session-info}
sessionInfo()
```
