---
title: "Visualisation"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 3
vignette: >
    %\VignetteIndexEntry{Visualisation}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

# Introduction

`ramen` provides `plot()` methods for all three core classes:
`ConsortiumMetabolism`, `ConsortiumMetabolismSet`, and
`ConsortiumMetabolismAlignment`. This vignette demonstrates each plot
type with synthetic and example data.

```{r load-library, message = FALSE}
library(ramen)
```

# Preparing example data

We create four synthetic consortia, combine them into a CMS, and
compute both pairwise and multiple alignments.

```{r create-data, message = FALSE}
cm_alpha <- synCM("Alpha", n_species = 5, max_met = 10, seed = 42)
cm_beta <- synCM("Beta", n_species = 5, max_met = 10, seed = 43)
cm_gamma <- synCM("Gamma", n_species = 4, max_met = 8, seed = 44)
cm_delta <- synCM("Delta", n_species = 6, max_met = 10, seed = 45)

cms <- ConsortiumMetabolismSet(
    cm_alpha, cm_beta, cm_gamma, cm_delta,
    name = "Demo"
)

cma_pair <- align(cm_alpha, cm_beta)
cma_mult <- align(cms)
```

# ConsortiumMetabolism plots

`plot(CM)` renders a directed metabolic flow network using `igraph`.
Nodes are coloured by role: **lightblue** = source (only outgoing),
**salmon** = sink (only incoming), **yellow** = intermediate (both).
The `type` argument selects which assay matrix to use for edge
weights.

## Binary network

The default view shows the structure of the metabolic network where
any edge present has weight 1.

```{r plot-cm-binary, fig.cap = "Binary metabolic network for Alpha."}
plot(cm_alpha, type = "Binary")
```

## Number of species per edge

Edge weights reflect how many species catalyse each pathway.

```{r plot-cm-nedges, fig.cap = "nEdges network for Alpha."}
plot(cm_alpha, type = "nEdges")
```

## Consumption and production

Consumption and production views weight edges by summed flux
magnitudes.

```{r plot-cm-consumption, fig.cap = "Consumption-weighted network."}
plot(cm_alpha, type = "Consumption")
```

```{r plot-cm-production, fig.cap = "Production-weighted network."}
plot(cm_alpha, type = "Production")
```

## Effective consumption and production

The effective diversity measures indicate how evenly species
contribute to each edge (Shannon-based effective number of species).

```{r plot-cm-effcons, fig.cap = "Effective consumption network."}
plot(cm_alpha, type = "EffectiveConsumption")
```

```{r plot-cm-effprod, fig.cap = "Effective production network."}
plot(cm_alpha, type = "EffectiveProduction")
```

# ConsortiumMetabolismSet plots

## Dendrogram

`plot(CMS)` draws the hierarchical clustering dendrogram with
numbered internal nodes. These node IDs are used by
`extractCluster()` to pull sub-clusters.

```{r plot-cms-dend, fig.cap = "CMS dendrogram with numbered nodes."}
plot(cms)
```

## Dendrogram with custom label colours

You can supply a tibble mapping leaf labels to colours.

```{r plot-cms-colours, fig.cap = "CMS dendrogram with coloured labels."}
colour_map <- data.frame(
    label = c("Alpha", "Beta", "Gamma", "Delta"),
    colour = c("steelblue", "firebrick",
               "forestgreen", "darkorange")
)
plot(cms, label_colours = colour_map)
```

# ConsortiumMetabolismAlignment plots

The CMA `plot()` method supports three types: `"heatmap"`,
`"network"`, and `"scores"`. The default depends on the alignment
type: `"network"` for pairwise, `"heatmap"` for multiple.

## Heatmap (multiple alignment)

The heatmap shows pairwise FOS similarities with dendrogram-based
ordering. Values range from 0 (no overlap) to 1 (identical).

```{r plot-cma-heatmap, fig.cap = "Similarity heatmap."}
plot(cma_mult, type = "heatmap")
```

## Network (pairwise alignment)

The network view visualises the alignment result as a directed
metabolite flow graph. Edge colours indicate the source:
**green** = shared pathways, **blue** = unique to query (Alpha),
**red** = unique to reference (Beta).

```{r plot-cma-network, fig.cap = "Pairwise alignment network."}
plot(cma_pair, type = "network")
```

## Scores (pairwise)

The score bar chart displays all computed similarity metrics for a
pairwise alignment.

```{r plot-cma-scores-pair, fig.cap = "Pairwise alignment scores."}
plot(cma_pair, type = "scores")
```

## Scores (multiple alignment)

For a multiple alignment, the scores plot shows summary statistics
across all pairwise comparisons.

```{r plot-cma-scores-mult, fig.cap = "Multiple alignment scores."}
plot(cma_mult, type = "scores")
```

# Using `plotDirectedFlow()` directly

For fine-grained control over the network layout, you can call
`plotDirectedFlow()` directly on an igraph object. This is the
function underlying all CM and CMA network plots.

```{r plot-directed-flow, fig.cap = "Custom directed flow plot."}
g <- igraph::graph_from_adjacency_matrix(
    SummarizedExperiment::assay(cm_beta, "nEdges"),
    mode = "directed",
    weighted = TRUE
)

plotDirectedFlow(
    g,
    color_edges_by_weight = TRUE,
    edge_width_range = c(0.5, 2),
    vertex_size = 12,
    vertex_label_cex = 0.7,
    main = "Beta - nEdges (custom)"
)
```

# Summary of plot types

| Object class | `type` argument | Description |
|-------------|-----------------|-------------|
| CM | `"Binary"` (default) | Directed network, unweighted |
| CM | `"nEdges"` | Edges weighted by species count |
| CM | `"Consumption"` | Edges weighted by consumption flux |
| CM | `"Production"` | Edges weighted by production flux |
| CM | `"EffectiveConsumption"` | Effective consuming species |
| CM | `"EffectiveProduction"` | Effective producing species |
| CMS | (none) | Dendrogram with numbered nodes |
| CMA | `"heatmap"` | Pairwise similarity heatmap |
| CMA | `"network"` | Shared/unique pathway graph |
| CMA | `"scores"` | Bar chart of similarity metrics |

# Session info

```{r session-info}
sessionInfo()
```
